---
title: "Basics"
---

## 1. How to tabulate in R frequencies of a survey answers (weighted)

When exploring survey data I often like to have a look at the frequencies of answers to particular questions, broken down by respondent's country, sex, age or any other variable at hand. However, this task isn't that simple, as survey data are often coded in very inconvenient way, for instance: 

```{r example_data, eval=TRUE, include=TRUE, echo=FALSE}
library(knitr)
library(kableExtra)
data <- read.csv("survey_data.csv")
kable(data[100:105,], "html") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, 
                position = "left")
```

With other types of data you would normally render table of frequencies with sort of the following code:

```{r, eval =  FALSE, include = TRUE, echo = TRUE}
data %>% 
  count(Question) %>% 
  mutate(prop = prop.table(n))
```
However, the characteristic feature of surveys is that each observation is assigned a unique weight, so the sample of respondents reflects the structure of relevant population. One needs to multiply each variable by relevant weights to obtain correct averages, frequencies and other statisics. But how would you do that with character variable? Solution seems simple - create dummy variables for each level of character variable ("strongly", "agree", etc.) and multiply them by weights. However, coding this in R is a bit tedious.

Fortunately, there is an easier solution, namely "weights" package, which contains a function *wpct*, which was designed to do exactly that. Its syntax is quite straightforward. You just need to pass relevant column of data and specify column with weights. For more legibility you may want to define the output as dataframe and convert fractions to percentages:
```{r, eval = TRUE, include = TRUE, echo = TRUE}
freq_table <- data$Question %>% 
              wpct(data$WeightPerCountry) %>%
              as.data.frame()

freq_table$. <- freq_table$. %>% scales::percent()
freq_table
```
Now, let's complicate things a bit. I suppose we would like to see how these percentages differ for various subpopulations, in our example: men vs women. First let's save a list of answers to variable, for later use.  
```{r, eval = TRUE, include = TRUE, echo = TRUE}
answers <- names(wpct(data$Question, weight = data$WeightPerCountry))
```

We need to apply wpct to each answer and each group separately. Fortunately, with a little trick we can pass wpct into margritr pipes, so we don't need to resolve to loops. The trick is to use "do" function.   
```{r, eval = TRUE, include = TRUE, echo = TRUE}
freq_table <- data %>% 
  group_by(Gender) %>%
  do(as.data.frame(wpct(.$Question, .$WeightPerCountry, na.rm = TRUE))) %>%
  mutate(answers = answers)
```
Now let's do few finishing touches: (1) change ugly default name of second column, (2) round fractions to double-digits and (3) spread the table to make comparisons easier. 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
names(freq_table)[2] <- "frequency"
freq_table$frequency <- freq_table$frequency %>% 
  percent(accuracy = 1)

freq_table <- freq_table %>% 
  spread(key = answers, "frequency")
```
And voila...
```{r echo=FALSE, message=FALSE, warning=FALSE}
kable(freq_table, "html") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, 
                position = "left")
```

## 2. Improving your code by writing functions. 

Most of the surveys contain a number of similiar questions approaching a particular topic from different angles. If you'd like to tabulate them one by one, you'll find yourself repeating the same code over and over again. To make your code more clear and efficient it's worth to write your own functions that will wrap the most repetitive chunks of your code. 

Writing your own functions sounds very proffessional, but is really easy. 
1. You start with coming up with a name to your function - usually something self-explainable. 
2. Next you assign a "function" to this name, open a round bracket an state the arguments this function should take (the bracket should close automatically).
3. Next open a curly bracket and put a chunk of code that works, remembering of the arguments of the function.
4. At the end of the function put "return", open a round bracket and name the object the function should return. 

As an example I will create a function that build a table with weighted answers to a particular survey question grouped by Gender.

```{r, eval = TRUE, include = TRUE, echo = TRUE}
survey_tabler <- function(data, question){
  answers <- names(wpct(pull(data[question]), weight = data$WeightPerCountry))
  
  freq_table <- data %>% 
  group_by(Gender) %>%
  do(as.data.frame(wpct(pull(.[question]), .$WeightPerCountry, na.rm = TRUE))) %>%
  mutate(answers = answers)
  
  names(freq_table)[2] <- "frequency"
  freq_table$frequency <- freq_table$frequency %>% 
  percent(accuracy = 1)

  freq_table <- freq_table %>% 
  spread(key = answers, "frequency")
  
  return(freq_table)
}
```

Now that we have our function, let's test it. 

```{r, eval = TRUE, include = TRUE, echo = TRUE}

survey_tabler(data, "Question")

```

